Ensemble Transport Adaptive Importance Sampling

### 2019-01-30 ETAIS.tex
remove red and fix bib for archiving

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. {\red As applications of these methods increase in size and
  complexity, the need for efficient methods increases. In this
paper, we present a particle ensemble algorithm.} At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms MCMC methods
  {\red with equivalent proposal distributions} for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, {\red multinomial transformation (MT),} which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\red
    for
  which speed-ups of orders of magnitude are demonstrated. Likelihood
  evaluations of the ensemble could be computed in a distributed
  manner, suggesting that this methodology is a good candidate
  for parallel Bayesian computations.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  {\red ensemble, resampling}.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

{\red In this paper, we present a framework for ensemble
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble.} The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. We also detail how algorithmic
parameters can be quickly tuned into optimal regimes, with respect to
the effective sample size statistic. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models. The method that we present is specifically designed to
tackle the challenges of certain types of low-dimensional inverse
problem which have complex posterior structure, for example
multimodality or strong correlations, and expensive
likelihood evaluations. We will demonstrate that this methodology
allows us to sample from such a posterior distribution with greater
precision per likelihood evaluation than 
standard MCMC chains of the same type.

@@

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:ETAIS}
we present the general framework of the ETAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of ETAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:MT} we introduce the {\red multinomial transformation (MT)} algorithm which is a less accurate but
faster alternative to resamplers which solve the optimal transport
problem exactly. In Section~\ref{sec:consistency} we consider
consistency of the ETAIS algorithm. {\red In Section~\ref{sec:useful} we briefly look at one
particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
some numerical examples, before a brief conclusion and discussion in
Section \ref{Sec:Conc}.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior {\red density $\pi$, with an ensemble
of weighted samples,} $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. These weighted samples can then be resampled, using the ETPF or otherwise,
into a {\red set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
\begin{equation}\label{eqn:resampler}
	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot),
\end{equation}}
{\red where $\delta_{x}(\cdot)$ is the Dirac delta measure.}

@@

{\red These deficiencies of MCMC methods
motivated the development of the Ensemble Transport Adaptive Importance Sampler
(ETAIS).}
In the next section we will introduce the method in its most
general form.

@@

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels  e.g. Gaussians centred at the current state, as in RWMH. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. These samples
are not stored in order to characterise the posterior density, since
the resampling process {\red is not exact}. They are
simply needed in order to inform the mixture proposal distribution for the next
iteration of the algorithm.

@@

{\red The choice of resampling algorithm is important since its
  output is used to formulate the proposal distribution for the next
  iteration. A basic resampler may be cheap to implement, but the
  resulting sample, and in turn the next iteration's proposal distribution, may not be as representative of the target
  distribution. This leads to higher variances of the weights in the
  importance sampler, and therefore to slower convergence to the
  target density.}

We wish to sample states $x \in X$ from a posterior
probability distribution with density {\red $\pi$}. Since we have $M$ ensemble members, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^\top$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) \sim
\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
variance of the proposal.

@@

Care needs to be taken when choosing kernel(s) for the proposal distribution to
ensure that the overall mixture is absolutely continuous with respect to
the posterior distribution. {\red For example, in many inverse problems we
may be looking to find the value of certain physical parameters which
may be strictly positive, or even bounded on an interval. In
this case, proposal kernels $\nu$ should be picked which have the same
support as the target. We will see an example of an unknown parameter
being bounded in Section \ref{sec:mixture}, where
$\beta$-distributed proposals are made to ensure that the distribution
is supported on $[0,1]$.}

@@

\section{Consistency of ETAIS}\label{sec:consistency}
{\red As with all importance sampling schemes, we must have absolute
  continuity of target distribution with respect to the proposal
  distribution if we wish to achieve convergence. This can usually be
  ensured by picking the proposal kernels $\nu$ to be from the same
  distribution type as the prior distribution.}

@@

\section{{\red A useful property of the ETAIS algorithm for multimodal
  distributions } }\label{sec:useful}

@@

{\red The numerics contained herein compare the ETAIS approach that we
  have presented against Metropolis-Hastings methods with
  the same type of proposal as is used by the kernels within
  ETAIS. Since we propose that ETAIS is a potential framework for
  parallelised Bayesian computations, we compare ETAIS with $N$
  ensemble members against an independent ensemble of $N$
  Metropolis-Hastings chains. All implementations were
  computed in serial, but the results demonstrate a clear speed-up
  which could be further exploited through parallelisation.
  }

@@

\subsection{Automated variance tuning}
\label{sec:problem 1}
{\red In this first example, we target a simple 1D Gaussian distribution, in
order to show the approximate equivalence of optimising the variance
of the proposal kernels using the effective sample size, and the $L^2$
error, which of course is not available to us in practical
applications.}

@@

\subsection{Multimodal targets and the effect of resampler quality}
\label{sec:bimodal}
{\red
In this second example we investigate the behaviour of the ETAIS
algorithm when applied to a bimodal problem, given by a mixture of two
Gaussians,
\[\pi(x) = 0.2\pi_1(x) + 0.8\pi_2(x),\]
where $\pi_1$ is the density of a $\mathcal{N}\left (\begin{pmatrix} 1
    \\ 1 \end{pmatrix}, \begin{pmatrix} 0.1 & 0 
    \\ 0 & 0.1 \end{pmatrix} \right )$ random variable and $\pi_2$ is the
  density of a $\mathcal{N}\left (\begin{pmatrix} -5
    \\ -5 \end{pmatrix}, \begin{pmatrix} 2.75 & -2.25 
    \\ -2.25 & 2.75 \end{pmatrix} \right )$ random variable.}

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes samples to new modes as they are found. This means
that we expect the number of samples in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\red In particular, we will look at the effect
of using either the ETPF, MT or a standard bootstrap resampler.

@@

\subsubsection{Implementation}
\label{sec:mixture_implementation}
{\red In order to quantify error, we note that} the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

Since the probability $p$ is constrained to lie in the interval
$[0,1]$, and the variances must be positive, it can be wasteful to use
Gaussian proposal distributions, {\red which will produce samples outside of
the support of the posterior. Moreover, the value of the variances of
the unknown distributions are strictly non-negative. The algorithm
will be most efficient if the proposal and posterior distributions are
mutually absolutely continuous. It is also useful to be able to pick
proposal densities for which the variance is easily scaled so that we
can tune them to optimise efficiency. Thus we pick the following
proposal distributions for the $p$, $\mu$ and $\sigma^2$ parameters, respectively;}
\[
	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
\]
where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* =
\sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be
tuned. This means that our proposal distributions will not be a
mixture of multivariate Gaussians, but independent mixtures of
univariate Beta, Gamma and Gaussian distributions.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.4^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.4^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.4^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

@@

Figure~\ref{fig:Lorenz_posterior} shows the marginal distributions of
the posterior distribution for the initial condition $\mathbf{x}_0$ as
found using the MH algorithm. While the $z$-dimension is largely
uncorrelated with the other two dimensions, the correlation between
$x$ and $y$ is {\red -0.97}, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.

@@

For this example, we use an ensemble size of {\red $M=1500$}, and the MT resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (MT). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Here the ETAIS results were produced with a serial
implementation, so this cost may be slightly overoptimistic, since the
runtimes do not include extra overheads of communicating the ensemble
states between processors. However, putting this together, we arrive at a speed-up factor of
over 430, which will undoubtedly overshadow any such underestimate of
the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one. 
}

@@

{\red We have explored the application of 
low dimensional Bayesian inverse problems. We have demonstrated numerically
that this method converges faster than the analogous naively parallelised
Metropolis-Hastings algorithms.} Further experimentation with the Metropolis
Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.

{\red The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger
ensembles. We have detailed how scaling parameters in the MCMC
proposals that are used within the mixture distribution can be quickly
and efficiently
tuned in an automated way. Lastly we have demonstrated that when the
likelihood is expensive, for instance because it involves the
numerical approximation of the solution of a differential equation, {\red
that we can achieve orders of magnitude reductions in cost to reach a
given error tolerance in comparison with standard Metropolis-Hastings approaches.}

However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there are curved
ridges in the density. {\red With increased dimensionality and/or
  complexity of the target the distribution, we also require increases
  in the size of the ensemble if we wish to have a stable ETAIS
  implementation. After a certain point, the extra overheads
  associated with a very large ensemble will outweigh the advantages
  of this approach. However, there} are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable ETAIS sampling {\red with much smaller ensemble sizes}.

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
%\newcommand{\revised}{\red}
\newcommand{\revised}{}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. {\revised As applications of these methods increase in size and
  complexity, the need for efficient methods increases. In this
paper, we present a particle ensemble algorithm.} At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms MCMC methods
  {\revised with equivalent proposal distributions} for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, {\revised multinomial transformation (MT),} which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\revised
    for
  which speed-ups of orders of magnitude are demonstrated. Likelihood
  evaluations of the ensemble could be computed in a distributed
  manner, suggesting that this methodology is a good candidate
  for parallel Bayesian computations.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  {\revised ensemble, resampling}.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

{\revised In this paper, we present a framework for ensemble
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
distribution from the current
ensemble.} The method makes use of a resampler based on optimal
transport which has been used in the context of particle
filters\cite{reich2013nonparametric}. We also detail how algorithmic
parameters can be quickly tuned into optimal regimes, with respect to
the effective sample size statistic. In particular we demonstrate the
advantages of this method when attempting to sample from multimodal
posterior distributions, such as those arising from inference for
mixture models. The method that we present is specifically designed to
tackle the challenges of certain types of low-dimensional inverse
problem which have complex posterior structure, for example
multimodality or strong correlations, and expensive
likelihood evaluations. We will demonstrate that this methodology
allows us to sample from such a posterior distribution with greater
precision per likelihood evaluation than 
standard MCMC chains of the same type.

@@

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:ETAIS}
we present the general framework of the ETAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of ETAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:MT} we introduce the {\revised multinomial transformation (MT)} algorithm which is a less accurate but
faster alternative to resamplers which solve the optimal transport
problem exactly. In Section~\ref{sec:consistency} we consider
consistency of the ETAIS algorithm. {\revised In Section~\ref{sec:useful} we briefly look at one
particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
some numerical examples, before a brief conclusion and discussion in
Section \ref{Sec:Conc}.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior {\revised density $\pi$, with an ensemble
of weighted samples,} $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. These weighted samples can then be resampled, using the ETPF or otherwise,
into a {\revised set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
\begin{equation}\label{eqn:resampler}
	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot),
\end{equation}}
{\revised where $\delta_{x}(\cdot)$ is the Dirac delta measure.}

@@

{\revised These deficiencies of MCMC methods
motivated the development of the Ensemble Transport Adaptive Importance Sampler
(ETAIS).}
In the next section we will introduce the method in its most
general form.

@@

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels  e.g. Gaussians centred at the current state, as in RWMH. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. These samples
are not stored in order to characterise the posterior density, since
the resampling process {\revised is not exact}. They are
simply needed in order to inform the mixture proposal distribution for the next
iteration of the algorithm.

@@

{\revised The choice of resampling algorithm is important since its
  output is used to formulate the proposal distribution for the next
  iteration. A basic resampler may be cheap to implement, but the
  resulting sample, and in turn the next iteration's proposal distribution, may not be as representative of the target
  distribution. This leads to higher variances of the weights in the
  importance sampler, and therefore to slower convergence to the
  target density.}

We wish to sample states $x \in X$ from a posterior
probability distribution with density {\revised $\pi$}. Since we have $M$ ensemble members, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^\top$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) \sim
\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
variance of the proposal.

@@

Care needs to be taken when choosing kernel(s) for the proposal distribution to
ensure that the overall mixture is absolutely continuous with respect to
the posterior distribution. {\revised For example, in many inverse problems we
may be looking to find the value of certain physical parameters which
may be strictly positive, or even bounded on an interval. In
this case, proposal kernels $\nu$ should be picked which have the same
support as the target. We will see an example of an unknown parameter
being bounded in Section \ref{sec:mixture}, where
$\beta$-distributed proposals are made to ensure that the distribution
is supported on $[0,1]$.}

@@

\section{Consistency of ETAIS}\label{sec:consistency}
{\revised As with all importance sampling schemes, we must have absolute
  continuity of target distribution with respect to the proposal
  distribution if we wish to achieve convergence. This can usually be
  ensured by picking the proposal kernels $\nu$ to be from the same
  distribution type as the prior distribution.}

@@

\section{{\revised A useful property of the ETAIS algorithm for multimodal
  distributions } }\label{sec:useful}

@@

{\revised The numerics contained herein compare the ETAIS approach that we
  have presented against Metropolis-Hastings methods with
  the same type of proposal as is used by the kernels within
  ETAIS. Since we propose that ETAIS is a potential framework for
  parallelised Bayesian computations, we compare ETAIS with $N$
  ensemble members against an independent ensemble of $N$
  Metropolis-Hastings chains. All implementations were
  computed in serial, but the results demonstrate a clear speed-up
  which could be further exploited through parallelisation.
  }

@@

\subsection{Automated variance tuning}
\label{sec:problem 1}
{\revised In this first example, we target a simple 1D Gaussian distribution, in
order to show the approximate equivalence of optimising the variance
of the proposal kernels using the effective sample size, and the $L^2$
error, which of course is not available to us in practical
applications.}

@@

\subsection{Multimodal targets and the effect of resampler quality}
\label{sec:bimodal}
{\revised
In this second example we investigate the behaviour of the ETAIS
algorithm when applied to a bimodal problem, given by a mixture of two
Gaussians,
\[\pi(x) = 0.2\pi_1(x) + 0.8\pi_2(x),\]
where $\pi_1$ is the density of a $\mathcal{N}\left (\begin{pmatrix} 1
    \\ 1 \end{pmatrix}, \begin{pmatrix} 0.1 & 0 
    \\ 0 & 0.1 \end{pmatrix} \right )$ random variable and $\pi_2$ is the
  density of a $\mathcal{N}\left (\begin{pmatrix} -5
    \\ -5 \end{pmatrix}, \begin{pmatrix} 2.75 & -2.25 
    \\ -2.25 & 2.75 \end{pmatrix} \right )$ random variable.}

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes samples to new modes as they are found. This means
that we expect the number of samples in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\revised In particular, we will look at the effect
of using either the ETPF, MT or a standard bootstrap resampler.

@@

\subsubsection{Implementation}
\label{sec:mixture_implementation}
{\revised In order to quantify error, we note that} the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

Since the probability $p$ is constrained to lie in the interval
$[0,1]$, and the variances must be positive, it can be wasteful to use
Gaussian proposal distributions, {\revised which will produce samples outside of
the support of the posterior. Moreover, the value of the variances of
the unknown distributions are strictly non-negative. The algorithm
will be most efficient if the proposal and posterior distributions are
mutually absolutely continuous. It is also useful to be able to pick
proposal densities for which the variance is easily scaled so that we
can tune them to optimise efficiency. Thus we pick the following
proposal distributions for the $p$, $\mu$ and $\sigma^2$ parameters, respectively;}
\[
	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
\]
where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* =
\sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be
tuned. This means that our proposal distributions will not be a
mixture of multivariate Gaussians, but independent mixtures of
univariate Beta, Gamma and Gaussian distributions.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution {\revised $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.4^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.4^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.4^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

@@

Figure~\ref{fig:Lorenz_posterior} shows the marginal distributions of
the posterior distribution for the initial condition $\mathbf{x}_0$ as
found using the MH algorithm. While the $z$-dimension is largely
uncorrelated with the other two dimensions, the correlation between
$x$ and $y$ is {\revised -0.97}, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.

@@

For this example, we use an ensemble size of {\revised $M=1500$}, and the MT resampling algorithm. The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations. Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.

@@

{\revised Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (MT). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Here the ETAIS results were produced with a serial
implementation, so this cost may be slightly overoptimistic, since the
runtimes do not include extra overheads of communicating the ensemble
states between processors. However, putting this together, we arrive at a speed-up factor of
over 430, which will undoubtedly overshadow any such underestimate of
the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one. 
}

@@

{\revised We have explored the application of 
low dimensional Bayesian inverse problems. We have demonstrated numerically
that this method converges faster than the analogous naively parallelised
Metropolis-Hastings algorithms.} Further experimentation with the Metropolis
Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.

{\revised The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger
ensembles. We have detailed how scaling parameters in the MCMC
proposals that are used within the mixture distribution can be quickly
and efficiently
tuned in an automated way. Lastly we have demonstrated that when the
likelihood is expensive, for instance because it involves the
numerical approximation of the solution of a differential equation, {\revised
that we can achieve orders of magnitude reductions in cost to reach a
given error tolerance in comparison with standard Metropolis-Hastings approaches.}

However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there are curved
ridges in the density. {\revised With increased dimensionality and/or
  complexity of the target the distribution, we also require increases
  in the size of the ensemble if we wish to have a stable ETAIS
  implementation. After a certain point, the extra overheads
  associated with a very large ensemble will outweigh the advantages
  of this approach. However, there} are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable ETAIS sampling {\revised with much smaller ensemble sizes}.
